From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting

### 2024-04-23 paper/paper.tex

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting},
  pdfauthor={Florian Börgel; Sven Karsten},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}}
\affiliation{1}{Leibniz-Institute for Baltic Sea Research Warnemünde, }
\correspondingauthor{Florian Börgel}{florian.boergel@io-warnemuende.de}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{runoff-data-used-for-training}{%
\subsection{Runoff data used for
training}\label{runoff-data-used-for-training}}

@@

\hypertarget{atmospheric-forcing}{%
\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}

@@

\hypertarget{ocean-model}{%
\subsection{Ocean Model}\label{ocean-model}}

@@

\hypertarget{lstm-network}{%
\subsection{LSTM network}\label{lstm-network}}

The Long Short-Term Memory (LSTM), a specialized form of Recurrent
Neural Networks (RNNs), is specifically tailored for modeling temporal
sequences. Its unique design allows it to adeptly handle long-range
dependencies, setting it apart from traditional RNNs in terms of
accuracy (see Figure~\ref{fig-lstm}).

@@

{\centering \includegraphics{draw_lstm.png}

@@

\end{figure}

This performance in modeling long-range dependencies has been validated
in various studies. The key component of LSTM's innovation is its memory
cell, \(c_t\)t, which stores state information, also refered to as
long-term memory. This cell is accessed, modified, and reset through
several self-parameterized gates. For the input of the sequence \(x_t\)
input, the forget gate \(f_t\) defines the percentage of the previous
long-term memory status \(c_{t-1}\) that should be retained
stored\hspace{0pt}. Next the input gate \(i_t\) decides how much of the
input is added to the the long-term memory, forming the updated cell
state \(c_{t}\). The decision to propagate the latest cell output,
\(c_t\), to the final state, \(h_t\), is governed by the output gate,
\(o_t\), representing the updated short-term memory of the hidden state
\(h_t\). A significant advantage of this architecture is the memory
cell's ability to retain gradients. This mechanism addresses the
vanishing gradient problem, where, as input sequences elongate, the
influence of initial stages becomes harder to capture, causing gradients
of early input points to approach zero. The LSTM's activation function,
inherently recurrent, mirrors the identity function with a consistent
derivative of 1.0, ensuring the gradient remains stable throughout
backpropagation.

One LSTM cell hence maybe expressed as:

\[
\begin{aligned}
i_t &= \sigma(W_{xi} x_t + W_{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i) \\
f_t &= \sigma(W_{xf} x_t + W_{hf} h_{t-1} + W_{cf} \circ c_{t-1} + b_f) \\
c_t &= f_t \circ c_{t-1} + i_t \circ \tanh(W_{xc} x_t + W_{hc} h_{t-1} + b_c) \\
o_t &= \sigma(W_{xo} x_t + W_{ho} h_{t-1} + W_{co} \circ c_t + b_o) \\
h_t &= o_t \circ \tanh(c_t)
\end{aligned}
\]

with

\begin{itemize}
\tightlist
\item
  \(x_t\): Input vector at time step \(t\).
\item
  \(h_{t-1}\): Hidden state from the previous time step.
\item
  \(C_{t-1}\): Cell state from the previous time step.
\item
  \(W\) and \(b\): Weight matrices and bias vectors, respectively,
  associated with the gates of the LSTM. The subscripts denote the
  specific gate or operation they are associated with (e.g., \(W_f\) and
  \(b_f\) are the weight matrix and bias for the forget gate).
\item
  \(\sigma\): Sigmoid activation function (\(0,\ldots 1\))
\item
  \(\tanh\): Hyperbolic tangent activation function (\$ -1 \ldots 1\$)
\end{itemize}

\hypertarget{convlstm-network}{%
\subsection{ConvLSTM network}\label{convlstm-network}}

The FC-LSTM fails to handle information when handling spatiotemporal
data due to its reliance on full connections in both input-to-state and
state-to-state transitions. To adress this limitation we use a convLSTM
architecture. convLSTM replaces the fully connected operations in the
LSTM with convolutional operations. Hence, all inputs
\(X_1, \ldots, X_t\), cell outputs \(C_1, \ldots, C_t\), hidden states
\(H_1, \ldots, H_t\), and gates \(i_t, f_t, o_t\) of the ConvLSTM are 3D
tensors. The last two dimensions of these tensors represent spatial
dimensions, specifically rows and columns. Conceptually, these inputs
and states can be visualized as vectors positioned on a spatial grid.

In the ConvLSTM, the future state of a specific cell on this grid is
determined by the inputs and past states of its neighboring cells. This
spatial consideration is integrated by employing a convolution operator
in both state-to-state and input-to-state transitions, as illustrated in
Fig. 2. The foundational equations for ConvLSTM are:

\[
\begin{aligned}
i_t &= \sigma(W_{xi} \ast X_t + W_{hi} \ast H_{t-1} + W_{ci} \circ C_{t-1} + b_i) \\
f_t &= \sigma(W_{xf} \ast X_t + W_{hf} \ast H_{t-1} + W_{cf} \circ C_{t-1} + b_f) \\
C_t &= f_t \circ C_{t-1} + i_t \circ \tanh(W_{xc} \ast X_t + W_{hc} \ast H_{t-1} + b_c) \\
o_t &= \sigma(W_{xo} \ast X_t + W_{ho} \ast H_{t-1} + W_{co} \circ C_t + b_o) \\
H_t &= o_t \circ \tanh(C_t)
\end{aligned}
\]

@@

\hypertarget{implemented-model-architecture}{%
\subsection{Implemented model
architecture}\label{implemented-model-architecture}}

@@

{\centering \includegraphics{model_structure.png}

@@

\end{figure}

@@

\hypertarget{tbl-letters}{}
\begin{longtable}[]{@{}ll@{}}
\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
\toprule()
Parameter name & Parameter size \\
\midrule()
\endfirsthead
\toprule()
Parameter name & Parameter size \\
\midrule()
\endhead
Channel size & 4 \\
Num. hidden layer & 10 \\
Num. timesteps & 30 \\
Conv. Kernelsize & (9,9) \\
Num. ConvLSTM layers & 1 \\
Batch size & 64 \\
~Learning Rate & 1e-3 with CosineAnnealing \\
\bottomrule()
\end{longtable}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

Figure~\ref{fig-PerformanceNeuralNetworkRunoff} shows the perfomance of
the model using the test dataset. The predicted total river runoff for
the Baltic Sea is closely matching the original data. Zooming in on the
largest individual rivers (lower panels) it can be seen that that also
the prediction of the inidivual rivers is close to the original data.

{\centering \includegraphics{images/paste-1.png}

\caption{\label{fig-PerformanceNeuralNetworkRunoff}Figure ..}

\end{figure}

The accuracy of the model is further displayed in
Figure~\ref{fig-statistical-evaluationNN} . The correlation is close to
1 and the residuals show the error of the model is below 1\% of the
original data.

{\centering \includegraphics{images/paste-4.png}

\caption{\label{fig-statistical-evaluationNN}Figure \ldots{}}

\end{figure}

@@

\includegraphics{images/paste-5.png}

\includegraphics{images/paste-6.png}

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

@@

\hypertarget{open-research}{%
\section{Open research}\label{open-research}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\leavevmode\vadjust pre{\hypertarget{ref-dee2011}{}}%
Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
and performance of the data assimilation system. \emph{Quarterly Journal
of the Royal Meteorological Society}, \emph{137}(656), 553--597.
\url{https://doi.org/10.1002/qj.828}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting},
  pdfauthor={Florian Börgel; Sven Karsten; Karoline Rummel},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}, Karoline
Rummel\affil{1}}
\affiliation{1}{Leibniz-Institute for Baltic Sea Research Warnemünde, }
\correspondingauthor{Florian Börgel}{florian.boergel@io-warnemuende.de}

@@

\section{Introduction}\label{introduction}

@@

\section{Methods}\label{methods}

\subsection{Runoff data used for
training}\label{runoff-data-used-for-training}

@@

\subsection{Atmospheric Forcing}\label{atmospheric-forcing}

@@

\subsection{Ocean Model}\label{ocean-model}

@@

\subsection{LSTM network}\label{lstm-network}

Before turning to the convolutional LSTM, the simpler architecture of
the plain LSTM is examined and should serve as a role model for the
later consideration of the full ConvLSTM.

The Long Short-Term Memory (LSTM), a specialized form of Recurrent
Neural Networks (RNNs), is specifically tailored for modeling temporal
sequences
\(\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]\) of
\(N_\tau\) input quantities
\(\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}\). The sequence
is taken from a dataset given in form of a time series \(\{\vec{X}^t\}\)
and the endpoint should coincide with the specific element in the time
series connected to time \(t\),
i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\). The \(N_k\) is the number
of used input ``channels'' and can, for example, represent different
measurable quantities. Its unique design allows it to adeptly handle
long-range dependencies, setting it apart from traditional RNNs in terms
of accuracy (see Figure~\ref{fig-lstm}).

@@

\centering{

\includegraphics{draw_lstm.png}

@@

\end{figure}%

This performance in modeling long-range dependencies has been validated
in various studies. The key component of LSTM's innovation is its memory
cell, \(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\), which
stores state information, also referred to as long-term memory. The
index \(\tau \in [1, N_\tau]\) corresponds to the \(\tau\)-th element in
the input sequence. The cell state is a vector consists of elements,
each associated with one of the \(N_h\) hidden layers, labeled by \(h\).
This vector is determined through several self-parameterized gates, all
in the same vector space as \(\vec{C}^t[\tau]\). The forget gate
\(\vec{F}^t[\tau]\) defines the percentage of the previous long-term
memory status \(\vec{C}^t[\tau-1]\) that should be retained
stored\hspace{0pt}. The input gate \(\vec{I}^t[\tau]\) decides how much
of the input is added to the the long-term memory, forming the updated
cell state. The so-called output gate \(\vec{O}^t[\tau]\) then
determines how much of the latest cell output, \(\vec{C}^t[\tau]\), is
propagated to the final state, \(\vec{H}^t[\tau]\) representing the
updated short-term memory of the hidden state.

For a fixed point \(\tau\) in the sequence, the action of a LSTM cell,
i.e.~the connection between the input \(\vec{X}^t[\tau]\), the various
gates and the state vectors, is given as follows. First, the elements of
the input sequence together with the hidden state are mapped onto
auxiliary gate vectors, collectively denoted by
\(\vec{g}^t[\tau] = (g^t_h[\tau]) \in \mathbb{R}^{N_h}\), via \[
\begin{aligned}
g_h^t[\tau] =  \mathcal{M}^{g}_{hk} \, X^t_k[\tau] +  \mathcal{N}^{g}_{hh'} \, H^t_{h'}[\tau-1] + \mathcal{B}^g_{h} \ ,
\end{aligned}
\]

where \(g=i,f,o,c\) stands for the input, forget, output and cell-state
gate, respectively and Eistein's summation convention is employed,
i.e.~indices that appear twice are summed over. The calligraphic symbols
\(\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}\) and
\(\mathcal{B}^g_{h}\) are the free parameters of the network that are
optimized for the given problem during the training, which is at the
heart of the any machine learning approach. The matrix
\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}\)
can be interpreted as a Markovian-like contribution of the current input
\(\vec{X}^t[\tau]\) to the gates, whereas the
\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}\)
scales a non-Markovian part determined by the hidden state of the last
sequence point \(\tau-1\). The vector
\(\vec{\mathcal{B}}^g = (\mathcal{B}^g_{h}) \in \mathbb{R}^{N_h}\) is a
learnable bias. It should be stressed that these parameters do neither
depend on \(t\) nor on \(\tau\) and are thus optimized once for the
complete dataset \(\{\vec{X}^t\}\).

Note that this mapping is sometimes extended by a contribution to the
\(g_h^t[\tau]\) from the past cell state \(\vec{C}^t[\tau-1]\)
(\textbf{XXX?}). Nevertheless, this mechanism called ``peeping'' is not
further considered in this work.

For the sake of brevity, we write the mapping more compactly in
matrix-vector form as
\begin{equation}\phantomsection\label{eq-LSTM-gates}{
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
}\end{equation}

Second, the actual gate vectors are computed by the core equations of
the LSTM as proposed by (\textbf{XXX?}):

\begin{equation}\phantomsection\label{eq-LSTM}{
\begin{aligned}
\vec{I}^t[\tau] &= \sigma(\vec{i}^t[\tau]) \\
\vec{F}^t[\tau] &= \sigma(\vec{f}^t[\tau]) \\
\vec{O}^t[\tau] &= \sigma(\vec{o}^t[\tau]) \\
\vec{C}^t[\tau] &= \vec{F}^t[\tau] \circ \vec{C}^t[\tau-1]  + \vec{I}^t[\tau] \circ \tanh(\vec{c}^t[\tau]) \\
\vec{H}^t[\tau]  &= \vec{O}^t[\tau] \circ \tanh(\vec{C}^t[\tau]) \ ,
\end{aligned}
}\end{equation}

where \(\sigma\) denotes the logisitc sigmoid function, \(\tanh\) is the
hyperbolic tangent and the \(\circ\) stands for the Hadamard product
(all applied in an element-wise fashion to the vectors). In the last two
equations the role of the input, forget and output gates as described
above become apparent.

The third step in a single layer LSTM (as employed for the work
presented here) is then to provide the output of the current LSTM cell,
i.e.~\(\vec{H}^t[\tau]\) and \(\vec{C}^t[\tau]\), to the subsequent LSTM
cell that processes the next element \(\vec{X}^t[\tau+1]\) of the input
sequence.

The full action of the LSTM network up to the end of the sequence can
therefore be written as a nested function call

\[
\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0] \right) \ldots \right) \right) \ ,
\]

where \(L\) represents Eqs. (Equation~\ref{eq-LSTM-gates},
Equation~\ref{eq-LSTM}) and the initial conditions are usually chosen as
\(\vec{H}^t[0]=\vec{C}^t[0]=0\).

The final output \(\vec{H}^t[N_\tau]\) and \(\vec{C}^t[N_\tau]\) encode
information on the full input sequence ending at time \(t\). This
information has to be decoded to obtain usueful information via an
appropriate subsequent layer, as it is described in the Sec.
(\textbf{XXX?}).

A significant advantage of this architecture is the memory cell's
ability to retain gradients. This mechanism addresses the vanishing
gradient problem, where, as input sequences elongate, the influence of
initial stages becomes harder to capture, causing gradients of early
input points to approach zero. The LSTM's activation function,
inherently recurrent, mirrors the identity function with a consistent
derivative of 1.0, ensuring the gradient remains stable throughout
backpropagation.

\subsection{ConvLSTM network}\label{convlstm-network}

Although the plain LSTM has high performance in handling temporal
sequences of point-like quantities it is not designed to recognize
spatial features in seuqences of two-dimensional maps. To adress this
limitation we use a convLSTM architecture.

In this kind of network the elements of the input sequence are given as
(kind of) vector fields
\(\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}\),
where \(x\in[1, N_x]\) and \(y \in [1, N_y]\) run over the horizontal
and vertical dimensions of the map, respectively. In order to enable the
``learning'' of spatial patterns, the free parameters of the network are
replaced by two-dimenmsional convolutional kernels
\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}\)
and
\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}\).
The width and the height of the kernels are given by \(N_\xi\) and
\(N_\eta\), respectively and
\(\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]\),
where we assume odd numbers for the kernel sizes.

The mapping from the input quantities to the gates is then given by a
convolution with these kernels \[
\begin{aligned}
g^t_h[x,y,\tau] & =  \mathcal{M}^{g}_{hk} [\xi,\eta]\, X^t_k[x-\xi, y-\eta, \tau]  +  \mathcal{N}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, \tau-1] + \mathcal{B}^g_{h}[x,y] \ .
\end{aligned}
\]

It becomes immediatly apparent that in case of the convLSTM, the bias,
gate and state vectors must become vector fields
(\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
this mapping in the same fashion as Eq. (Equation~\ref{eq-LSTM-gates})
but with replacing the normal matrix-vector multiplication by a
convolution (denoted with \(\ast\)), i.e.\\
\[
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
\]

The susebsequent processing of the \(\vec{g}^t[\tau]\) remains
symbolically the same as presented in Eq. (Equation~\ref{eq-LSTM}) but
with all appearing quantities meaning standing now for vector fields
instead of simple vectors.

@@

\subsection{Implemented model
architecture}\label{implemented-model-architecture}

@@

\centering{

\includegraphics{model_structure.png}

@@

\end{figure}%

@@

\begin{longtable}[]{@{}ll@{}}
\caption{Hyperparameters}\label{tbl-letters}\tabularnewline
\toprule\noalign{}
Parameter name & Parameter size \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter name & Parameter size \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Channel size & 4 \\
Num. hidden layer & 10 \\
Num. timesteps & 30 \\
Conv. Kernelsize & (9,9) \\
Num. ConvLSTM layers & 1 \\
Batch size & 64 \\
~Learning Rate & 1e-3 with CosineAnnealing \\
\end{longtable}

@@

\section{Results}\label{results}

@@

The accuracy of the model is further displayed in
\textbf{?@fig-statistical-evaluationNN} . The correlation is close to 1
and the residuals show the error of the model is below 1\% of the
original data.

\includegraphics{images/paste-13.png}

\textbf{?@fig-PerformanceNeuralNetworkRunoff} shows the performance of
the model using the test dataset. The predicted total river runoff for
the Baltic Sea is closely matching the original data. Zooming in on the
largest individual rivers (lower panels) it can be seen that that also
the prediction of the inidivual rivers is close to the original data.

\includegraphics{images/paper_total_river_runoff.png}

@@

\includegraphics{images/paste-9.png}

\includegraphics{images/paste-12.png}

\section{Acknowledgments}\label{acknowledgments}

@@

\section{Open research}\label{open-research}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-dee2011}
Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
and performance of the data assimilation system. \emph{Quarterly Journal
of the Royal Meteorological Society}, \emph{137}(656), 553--597.
\url{https://doi.org/10.1002/qj.828}

\bibitem[\citeproctext]{ref-fangExaminingApplicabilityDifferent2019}
Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
(2019). Examining the applicability of different sampling techniques in
the development of decomposition-based streamflow forecasting models.
\emph{Journal of Hydrology}, \emph{568}, 534--550.
\url{https://doi.org/10.1016/j.jhydrol.2018.11.020}
